---
title: "020719 Final Statistical Analysis"
author: "Kelly Ireland"
date: "3/2/2020"
output:
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Abbreviations used:
exp - experiment
dpf - days post fertilization
snv - snout-vent length
ed - eye diameter
sbl - swim bladder length
sba - swim bladder area

```{r}
exp <- read.csv("data020719.csv")

exp14 <- exp[which(exp$dpf=="14"),]
exp28 <- exp[which(exp$dpf=="28"),]

#oil
oil14 <- exp14[which(exp14$treatment=="oil"),]
oil28 <- exp28[which(exp28$treatment=="oil"),]
oil <- exp[which(exp$treatment=="oil"),]

#control
con14 <- exp14[which(exp14$treatment=="control"),]
con28 <- exp28[which(exp28$treatment=="control"),]
control <- exp[which(exp$treatment=="control"),]

str(exp)
exp$dpf <- as.factor(exp$dpf)

library(ggplot2)
library(nlme)
library(predictmeans)
library(effsize) #calculates cohen d
library(EMAtools) #calculates cohen d for each effect in nlme
```

```{r Summary Stats}
summary <- matrix(nrow=24, ncol=2, dimnames=list(c("OIL SVL 14 DPF", "CONTROL SVL 14", "OIL SVL 28 DPF", "CONTROL SVL 28 DPF", "OIL ED 14 DPF", "CONTROL ED 14 DPF","OIL ED 28 DPF", "CONTROL ED 28 DPF", "OIL SBL 14 DPF", "CONTROL SBL 14 DPF","OIL SBL 28 DPF", "CONTROL SBL 28 DPF","OIL SBA 14 DPF", "CONTROL SBA 14 DPF","OIL SBA 28 DPF","CONTROL SBA 28 DPF", "OIL IL-1B 14 DPF", "CONTROL IL-1B 14 DPF", "OIL IL-1B 28 DPF", "CONTROL IL-1B 28 DPF", "OIL TNF-A 14 DPF", "CONTROL TNF-A 14 DPF", "OIL TNF-A 28 DPF", "CONTROL TNF-A 28 DPF"), c("mean","sd")))
summary

#SNV
summary[1,1] <- mean(oil14$snv) #Oil treatment at 14 DPF
summary[1,2] <- sd(oil14$snv) #Oil treatment at 14 DPF
summary[2,1] <- mean(con14$snv) #Control treatment at 14 DPF
summary[2,2] <- sd(con14$snv) #Control treatment at 14 DPF
summary[3,1] <- mean(oil28$snv) #Oil treatment at 28 DPF
summary[3,2] <- sd(oil28$snv) #Oil treatment at 28 DPF
summary[4,1] <- mean(con28$snv) #Control treatment at 28 DPF
summary[4,2] <- sd(con28$snv) #Control treatment at 28 DPF
#ED
summary[5,1] <- mean(oil14$ed)
summary[5,2] <- sd(oil14$ed)
summary[6,1] <- mean(con14$ed)
summary[6,2] <- sd(con14$ed)
summary[7,1] <- mean(oil28$ed)
summary[7,2] <- sd(oil28$ed)
summary[8,1] <- mean(con28$ed)
summary[8,2] <- sd(con28$ed)
#SBL
summary[9,1] <- mean(oil14$sbl) #Oil treatment at 14 DPF
summary[9,2] <- sd(oil14$sbl) #Oil treatment at 14 DPF
summary[10,1] <- mean(con14$sbl) #Control treatment at 14 DPF
summary[10,2] <- sd(con14$sbl) #Control treatment at 14 DPF
summary[11,1] <- mean(oil28$sbl) #Oil treatment at 28 DPF
summary[11,2] <- sd(oil28$sbl) #Oil treatment at 28 DPF
summary[12,1] <- mean(con28$sbl) #Control treatment at 28 DPF
summary[12,2] <- sd(con28$sbl) #Control treatment at 28 DPF
#SBA
summary[13,1] <- mean(oil14$sba) #Oil treatment at 14 DPF
summary[13,2] <- sd(oil14$sba) #Oil treatment at 14 DPF
summary[14,1] <- mean(con14$sba) #Control treatment at 14 DPF
summary[14,2] <- sd(con14$sba) #Control treatment at 14 DPF
summary[15,1] <- mean(oil28$sba) #Oil treatment at 28 DPF
summary[15,2] <- sd(oil28$sba) #Oil treatment at 28 DPF
summary[16,1] <- mean(con28$sba) #Control treatment at 28 DPF
summary[16,2] <- sd(con28$sba) #Control treatment at 28 DPF
#IL1B
summary[17,1] <- mean(oil14$il1brelative_normalized_expression,na.rm=TRUE) #oil 14
summary[17,2] <- sd(oil14$il1brelative_normalized_expression,na.rm=TRUE) #oil 14
summary[18,1] <- mean(con14$il1brelative_normalized_expression,na.rm=TRUE) #con 14
summary[18,2] <- sd(con14$il1brelative_normalized_expression,na.rm=TRUE) #con14
summary[19,1] <- mean(oil28$il1brelative_normalized_expression,na.rm=TRUE) #oil 28
summary[19,2] <- sd(oil28$il1brelative_normalized_expression,na.rm=TRUE) #oil 28
summary[20,1] <- mean(con28$il1brelative_normalized_expression,na.rm=TRUE) #con 28
summary[20,2] <- sd(con28$il1brelative_normalized_expression,na.rm=TRUE) #con 28
#TNF
summary[21,1] <- mean(oil14$tnfrelative_normalized_expression,na.rm=TRUE) #oil 14
summary[21,2] <- sd(oil14$tnfrelative_normalized_expression,na.rm=TRUE) #oil 14
summary[22,1] <- mean(con14$tnfrelative_normalized_expression,na.rm=TRUE) #con 14
summary[22,2] <- sd(con14$tnfrelative_normalized_expression,na.rm=TRUE) #con14
summary[23,1] <- mean(oil28$tnfrelative_normalized_expression,na.rm=TRUE) #oil 28
summary[23,2] <- sd(oil28$tnfrelative_normalized_expression,na.rm=TRUE) #oil 28
summary[24,1] <- mean(con28$tnfrelative_normalized_expression,na.rm=TRUE) #con 28
summary[24,2] <- sd(con28$tnfrelative_normalized_expression,na.rm=TRUE) #con 28

?mean
summary
rdsum <- round(summary, digits=3)
rdsum
```


```{r Snout Vent Length}
snvm1 <- lme(snv ~ treatment + dpf, random =~1 | flask, weights=varIdent(form=~1 | dpf), data = exp)
summary(snvm1)
plot(snvm1)
plot(snvm1,resid(.)~fitted(.))
qqnorm(snvm1)
str(exp)

snvm2 <- lme(snv ~ treatment * dpf, random =~1 | flask, weights=varIdent(form=~1 | dpf), data = exp)
summary(snvm2)

lme.dscore(snvm1, data=exp, type ="nlme")

#significant negative effect of oil treatment on SNV
#significant positive effect of dpf on SNV

#pull out residuals from linear model and then compare two treatments using variance test 
#may have more variance in oil because development is complex process and oil could be affecting development in a variety of ways.
```
Treatment (negative relationship) and dpf (positive relationship) had significant effect on snv. 

```{r Eye Diameter}
edm1 <- lme(ed ~ treatment + dpf, random =~1 | flask, weights=varIdent(form=~1 | dpf), data = exp)
summary(edm1)
plot(edm1)
plot(edm1,resid(.)~fitted(.))

edm2 <- lme(ed ~ treatment * dpf, random =~1 | flask, weights=varIdent(form=~1 | dpf), data = exp)
summary(edm2)

#cohen-d effect size calculation, but what about considerations for flask, dpf, etc?
cohen.d(exp$ed,exp$treatment,pooled=TRUE,paired=FALSE,
na.rm=TRUE, mu=0, hedges.correction=FALSE,
conf.level=0.95,noncentral=FALSE,
within=TRUE, subject=NA)

lme.dscore(edm1, data=exp, type ="nlme")
```
Dpf has a significant positive effect on eye diameter. Oil treatment has a significant negative relationship on eye diameter. 

```{r correlation betweeen SVL and SBL?}
x <- exp$snv
y <- exp$sbl
cor(x,y,use="complete.obs",method="pearson")
#correlation not significant
```


```{r Swim Bladder Length}
sblm1 <- lme(sbl ~ treatment + dpf, random =~1 | flask, weights=varIdent(form=~1 | dpf), data = exp)
summary(sblm1)
plot(sblm1)
plot(sblm1,resid(.)~fitted(.))

sblm2 <- lme(sbl ~ treatment * dpf, random =~1 | flask, weights=varIdent(form=~1 | dpf), data = exp)
summary(sblm2)

bartlett.test(sbl ~ treatment, data = exp)
#no variance between treatments

lme.dscore(sblm1, data=exp, type ="nlme")
```
Dpf has a significant positive effect on swim bladder length. Treatment has almost no effect on SBL.

```{r Swim Bladder Area}
sbam1 <- lme(sba ~ treatment + dpf, random =~1 | flask, weights=varIdent(form=~1 | dpf), data = exp)
summary(sbam1)
plot(sbam1)
plot(sbam1,resid(.)~fitted(.))

sbam2 <- lme(sba ~ treatment * dpf, random =~1 | flask, weights=varIdent(form=~1 | dpf), data = exp)
summary(sbam2)

lme.dscore(sbam1, data=exp, type ="nlme")
```
Days post fertilization has a positive significant relationship with swim bladder area. Treatment doesn't have a significant effect on sba.  

```{r qrtpcr data read in}
qexp <- read.csv("020719qrtpcr.csv")
```


```{r IL1B expression}
il1bm1 <- lme(il1brelative_normalized_expression ~ treatment * dpf, random =~1 | flask, weights=varIdent(form=~1 | dpf), data = qexp)
summary(il1bm1)
plot(il1bm1)
plot(il1bm1,resid(.)~fitted(.))

il1bm2 <- lme(il1brelative_normalized_expression ~ treatment + dpf + initial_rna_conc, random =~1 | flask, weights=varIdent(form=~1 | dpf), data=qexp)
summary(il1bm2)


lme.dscore(il1bm1, data=qexp, type ="nlme")
```
Significant effect of dpf (positive relationship) and interaction between treatment and oil (negative relationship) on IL1B expression

```{r}
tnfm1 <- lme(tnfrelative_normalized_expression ~ treatment * dpf, random =~1 | flask, weights=varIdent(form=~1 | dpf), data = exp, na.action = na.omit)
summary(tnfm1)
plot(tnfm1)
plot(tnfm1,resid(.)~fitted(.))

tnfm2 <- lme(tnfrelative_normalized_expression ~ treatment + dpf + initial_rna_conc, random =~1 | flask, weights=varIdent(form=~1 | dpf), data=qexp)
summary(tnfm2)
#initial rna concentration does not have significant effect on the expression of tnf

lme.dscore(tnfm1, data=qexp, type ="nlme")
```
Significant effect of dpf (positive relationship) and interaction between treatment and dpf (negative relationship) of treatment and oil on IL1B expression. 

```{r separate variances by time? Yes}
bartlett.test(snv ~ dpf, data = exp)
bartlett.test(ed ~ dpf, data = exp)
bartlett.test(sbl ~ dpf, data = exp)
bartlett.test(sba ~ dpf, data = exp)
bartlett.test(il1brelative_normalized_expression ~ dpf, data = exp)
bartlett.test(tnfrelative_normalized_expression ~ dpf, data = exp)

#is there different variances between control and oil at 28 dpf?
bartlett.test(il1brelative_normalized_expression ~ treatment, data = exp14)
#highly significant variance at 14 dpf p=0.0007093
bartlett.test(il1brelative_normalized_expression ~ treatment, data = exp28)
#highly significant differnt variances p=1.659e-05

bartlett.test(tnfrelative_normalized_expression ~ treatment, data = exp14)
#significant different variances p=0.001333
bartlett.test(tnfrelative_normalized_expression ~ treatment, data = exp28)
#highly significant differnt variances p=2.91e-06
```


